---
title: "R Coding Exercise"
---

Placeholder file for the future R coding exercise.

This coding exercise begins with loading and checking the data using help(), str(), summary(), and class().

```{r}
# Loading and Checking Data

# loading dslabs package
library(dslabs)

# look at help file for gapminder data
help(gapminder)

# get an overview of data structure
str(gapminder)

# get a summary of the data
summary(gapminder)

# determine the type of object gapminder is
class(gapminder)
```

This part of the exercise is titled, "Processing Data", where objects are creating using the gapminder data set.

```{r}
# Processing Data

# load tidyverse package
library(tidyverse)

# create new object/variable called africadata
africadata <- gapminder %>%
  filter(continent == "Africa")

# create a new object on only infant_mortality
infant_health <- africadata %>%
  select(infant_mortality, life_expectancy)

# Create a new object containing population and life_expectancy
overall_health <- africadata %>%
  select(population, life_expectancy)

# inspect new objects
str(infant_health)
summary(infant_health)

str(overall_health)
summary(overall_health)
```

This part of the exercise is titled, "Plotting", and focuses on creating plots on the objects previously created. For the graph, "Life expectancy as a function of Population Size in Africa", there is a streaking pattern which is caused by countries having several entries in different years. Countries tended to increase in life expectancy and size as years pass.

```{r}
library(ggplot2)

# create a scatter plot of life expectancy as a function of infant mortality
infant_graph <- ggplot(africadata, aes(x = infant_mortality, y =life_expectancy)) + geom_point(alpha = 0.6, color = "blue") + labs(
  title = "Life expectancy as a function of Infant Mortality in Africa",
  x = "Infant Mortality",
  y = "Life Expectancy"
)

# display graph
print(infant_graph)

# create a scatter plot of life expectancy as a function of population size
population_graph <- ggplot(africadata, aes(x = population, y = life_expectancy)) + geom_point(alpha = 0.6, color = "blue") + scale_x_log10() + labs(
  title = "Life expectancy as a function of Population Size in Africa",
  x = "Population (Log Scale)",
  y = "Life Expectancy"
)

# display graph
print(population_graph)

```

This part of the exercise is titled, "More data processing". In this part, the data is inspected to figure out which years have the most missing counts. The assignment chooses the year 2000 to filter the data. I create an object called africadata_twothousand which contains the filtered data.

```{r}
na_count_per_year <- africadata %>%
  filter(is.na(infant_mortality)) %>%
  group_by(year) %>%
  summarize(na_count_per_year = n())

# print result
print(na_count_per_year)

# create new object with only the year 2000
africadata_twothousand <- africadata %>%
  filter(year == "2000")

# check/inspect new object with data from the year 2000
str(africadata_twothousand)
summary(africadata_twothousand)
```

This part of the exercise is called, "More plotting". This is essentially identidical as a previous section but utilizes filtered data for only the year 2000.

```{r}
# create a scatter plot of life expectancy as a function of infant mortality
infant_graph_twothousand <- ggplot(africadata_twothousand, aes(x = infant_mortality, y =life_expectancy)) + geom_point(alpha = 0.6, color = "blue") + labs(
  title = "Life expectancy as a function of Infant Mortality in Africa in the year 2000",
  x = "Infant Mortality",
  y = "Life Expectancy"
)

# display graph
print(infant_graph_twothousand)

# create a scatter plot of life expectancy as a function of population size
population_graph_twothousand <- ggplot(africadata_twothousand, aes(x = population, y = life_expectancy)) + geom_point(alpha = 0.6, color = "blue") + scale_x_log10() + labs(
  title = "Life expectancy as a function of Population Size in Africa in the year 2000",
  x = "Population (Log Scale)",
  y = "Life Expectancy"
)

# display graph
print(population_graph_twothousand)

```

This part of the exercise is called, "Simple model fits". This code creates a linear model using lm(). Model 1, "Fit1", is a model with life expectancy as the outcome and infant mortality as the predictor. This model has a p-value of 2.826e-08. This indicates significance. Model 2, "Fit2", is a model with population size as a predictor and life expectancy as the outcome. This model has a p-value of 0.6159. This indicates insignificance.

```{r}
# Fit Model 1 - Life Expectancy and Infant Mortality
fit1 <- lm(life_expectancy ~ infant_mortality, data = africadata_twothousand)

# summary of fit model 1
summary(fit1)

# Fit Model 2 - Life Expectancy and Population Size
fit2 <- lm(life_expectancy ~ population, data = africadata_twothousand)

# summary of fit model 2
summary(fit2)
```

# The following sections contributed by Guozheng Yang

# Load packages and dataset

I'm going to explore the *murders* dataset from the package *dslabs*. This dataset contains gun murder data for 2010 reported by FBI, and is sourced from Wikipedia. The dataset is organized by states, with the population of each state included. 

```{r}
# Load required package
library(dslabs) # This package has the murders dataset
library(tidyverse)
library(ggplot2)
library(maps) # This package is used to make a map

# Look at help file for murders data
help(murders)
```

First of all, let's take a look at the data structure. 

```{r}
# Determine the type of murders
class(murders)

# Get an overview of data structure
str(murders)

# Get a summary of data
summary(murders)

# Check if there are missing values
anyNA(murders)
```

As shown, there are 51 observations and 5 variables in this dataset. 51 states and their abbreviations are listed. The states are also classified by their geographic locations. Two numeric variables denote the population and number of gun murders of each state in 2010. Luckily, we don't have any NA in this dataset. 

# Map plotting

Since we have data from different states, I want to make a map to compare the different gun murder rates by states. Here I calculate the gun murder rates as the number of gun murders divided by the population of each state. Then I use a heatmap to color the statesby their gun murder rates. Of note, the US map data is in the package *maps*, with the longitude and latitude of each state included. This tool makes it convenient for us to draw the heatmap. 

Let's take a look at the data from package *maps* first.

```{r}
# Extract the US map data
us_map <- map_data("state")

# Determine the type of murders
class(us_map)

# Get an overview of data structure
str(us_map)

# Get a summary of data
summary(us_map)
```

As shown, all state names are in stored in the *region* variable in lower case. Variables *long* and *lat* have the longitude and latitude of each state's border. As this is a well prepared data, I will calculate gun murder rates in the *murders* dataset and then merge it into *us_map*.

```{r}
# Prepare the murders data
murders <- murders %>%
  mutate(state_low=tolower(state), # Convert state names to lower case
         rate=total/population) # Calculate gun murder rate

map_data <- us_map %>%
  left_join(murders, by=c("region"="state_low"))
```

As the data for plotting is well prepared, I will use *geom_polygon()* function to draw the map and then color each state by gun murder rates. The *geom_polygon()* function is a useful tool for connecting a group of points by a certain order. In our case, it's used to draw the border line of each state.

```{r}
# Draw a heatmap of gun murder rates
us_gmr <- ggplot(map_data, aes(x=long, y=lat, group=group, fill=rate))+
  geom_polygon(color="steelblue4", linewidth=1)+ # Draw border line of states
  scale_fill_gradient(name="Gun murder rate", 
                      low="white", high="firebrick4")+ # Define color gradient
  theme_bw()+
  labs(title="Gun murder rate by state", x="Longitude", y="Latitude")+
  theme(axis.title.x=element_text(size=10, color="black", face="bold"),
        axis.title.y=element_text(size=10, color="black", face="bold"),
        axis.text.x=element_text(color="black", size=8),
        axis.text.y=element_text(color="black", size=8),
        plot.title=element_text(size=15, color="black", face="bold"),
        legend.position="right",
        legend.title=element_text(size=10, face="bold"))

us_gmr
```

As the map shows, the gun murder rate is higher overall in the southern US. For the northern US, the northwest has a lower gun murder rate than the northeast. This map demonstrate the geographic distribution of gun murder rates. 

# Model fitting

As the response of interest is gun murder rate, it's intuitive to fit a Poisson regression with a *log* link function. In the Poisson regression model, the number of gun murders is the response, and the population of each state is the offset. The region of states is a predictor to show if gun murder rate is different across different regions.

Before that, let's make a grouped boxplot to see if the difference is evident. Here I use *region* as the X-axis and gun murder rate as the Y-axis. 

```{r}
boxplot <- ggplot(murders, aes(x=region, y=rate))+
  geom_boxplot(color="firebrick3", width=.5, linewidth=1)+
  theme_bw()+
  labs(title="Gun murder rate by region", x="Region", y="Gun murder rate")+
  theme(axis.title.x=element_text(size=10, color="black", face="bold"),
        axis.title.y=element_text(size=10, color="black", face="bold"),
        axis.text.x=element_text(color="black", size=8),
        axis.text.y=element_text(color="black", size=8),
        plot.title=element_text(size=15, color="black", face="bold"),
        legend.position="right",
        legend.title=element_text(size=10, face="bold"))
boxplot
```

As shown, the gun murder rate in southern US is evidently higher compared to other regions. The northeast, north central region, and west of the US have similar gun murder rate, though the median in the west is a little lower.

Now, let's fit a Poisson regression model to see if the regional difference is statistically significant. As indicated by the map above, the southern US has a generally lower gun murder rate. So I want to use *South* as the reference group and adjust the factor levels of *region*.

```{r}
# Adjust factor level of region
murders$region_fct <- factor(murders$region, levels=c("South", "North Central", "West", "Northeast"))

# Poisson regression: gun murder rate ~ region
poi_fit <- glm(total ~ region_fct, offset=log(population), family="poisson", data=murders)
summary(poi_fit)

# Exponential the coefficient for interpretation
exp(poi_fit$coefficients)
```

According to the output, the mean gun murder rate in southern US is 3.63e-5. The north central US has a 25% decrease in gun murder rate compared to southern US. The western and northeastern US both have about 27% decrease in gun murder rate than southern US. Notably, the coefficient for north central US, western US, and northeastern US are all negative with p-values lower than 0.05. We have enough evidence to reject the null hypothesis and conclude that the gun murder rates in these three regions are significantly lower compared to southern US.

