---
title: "CDC Data Exercise"
author: "Vincent Nguyen"
date: "February 3, 2025"
---

# Coding Exercise for Week 5

## Package Loading

First, we will start with loading some necessary packages for data creation, visualization, and more.

```{r}
# loading dslabs package
library(dplyr)
library(purrr)
library(lubridate)
library(ggplot2)
library(here)
library(readr)
library(janitor)
```

## Data Set

The data set is titled, "[Weekly Provisional Counts of Deaths by State and Select Causes, 2020-2023](https://data.cdc.gov/NCHS/Weekly-Provisional-Counts-of-Deaths-by-State-and-S/muzy-jte6/about_data)". The data set contains 10476 observations of 35 variables. It covers counts of death by nationally or by state. Additionally, it includes causes of death and more.

```{r}
# Import dataset

# path to data
# note the use of the here() package and not absolute paths
data_location <- here::here("cdcdata-exercise", "Weekly_Provisional_Counts_of_Deaths_by_State_and_Select_Causes__2020-2023_20250204.csv")
data <- read.csv(data_location)%>%
  clean_names()

# filterd for only the US (no states) and then also removed some diseases not of interest
# also rename the columns because they are formatted weird
data_filtered <- data %>%
  filter(jurisdiction_of_occurrence == "United States") %>%
  select(1:7, 9:10, 12, 17) %>%
  rename(
    cancer_count = malignant_neoplasms_c00_c97,
    diabetes_count = diabetes_mellitus_e10_e14,
    influenza_pneumonia_count = influenza_and_pneumonia_j09_j18,
    heart_disease_count = diseases_of_heart_i00_i09_i11_i13_i20_i51
  )


```

## Summary Stats

Creation of summary statistics for the entire data set and for diseases of interest.

```{r}
# summary statistics of all causes of death
summary(data_filtered)

# mmean / sd of cancer deaths
mean(data_filtered$cancer_count)
sd(data_filtered$cancer_count)

# mean / sd of diabetes
mean(data_filtered$diabetes_count)
sd(data_filtered$diabetes_count)

# mean / sd of heart disease
mean(data_filtered$heart_disease_count)
sd(data_filtered$heart_disease_count)

# mean / sd of influenza and pneumonia
mean(data_filtered$influenza_pneumonia_count)
sd(data_filtered$influenza_pneumonia_count)

```

## Graphs for Distributions

Creation of graphs to visualize the distribution of death count cases on a MMWR week-to-week basis. Surprisingly, they look almost normal with some skewing on some of the graphs.

```{r}
# Distribution graph of cancer deaths
cancer_graph <- ggplot(data_filtered, aes(x = cancer_count)) +
  geom_histogram(binwidth = 50, fill = "skyblue", color = "black") +
  labs(
    x = "Cancer Death Count",
    y = "Frequency",
    title = "Distribution of Cancer Death Counts"
  )

print(cancer_graph)

# Distribution graph of diabetes deaths
diabetes_graph <- ggplot(data_filtered, aes(x = diabetes_count)) +
  geom_histogram(binwidth = 50, fill = "skyblue", color = "black") +
  labs(
    x = "Diabetes Death Count",
    y = "Frequency",
    title = "Distribution of Diabetes Death Counts"
  )

print(diabetes_graph)


# Distribution graph of heart disease deaths
heart_disease_graph <- ggplot(data_filtered, aes(x = heart_disease_count)) +
  geom_histogram(binwidth = 50, fill = "skyblue", color = "black") +
  labs(
    x = "Heart Disease Death Count",
    y = "Frequency",
    title = "Distribution of Heart Disease Death Counts"
  )

print(heart_disease_graph)

# Distribution graph of influenza/pneumonia deaths
influenza_pneumonia_graph <- ggplot(data_filtered, aes(x = influenza_pneumonia_count)) +
  geom_histogram(binwidth = 50, fill = "skyblue", color = "black") +
  labs(
    x = "Influenza/Pneumonia Death Count",
    y = "Frequency",
    title = "Distribution of Influenza/Pneumonia Death Counts"
  )

print(influenza_pneumonia_graph)

```

## Synthesizing New Data
This section was contributed by Connor Norris.

```{r}
#Additional exploratory analysis

#Mean and SD of all-cause mortality
paste("Mean of All-Cause Mortality:", mean(data_filtered$all_cause))
paste("SD of All-Cause Mortality:", sd(data_filtered$all_cause))

#Histogram of all-cause mortality
ggplot(data_filtered, aes(x = all_cause)) +
  geom_histogram(binwidth = 500, fill = "skyblue", color = "black") +
  labs(
    x = "All Cause Mortality Count",
    y = "Frequency",
    title = "Distribution of All Cause Mortality"
  )

#Mean and SD of natural cause mortality
paste("Mean of Natural Cause Mortality:", mean(data_filtered$natural_cause))
paste("SD of Natural Cause Mortality:", sd(data_filtered$natural_cause))

#Histogram of natural cause mortality
ggplot(data_filtered, aes(x = natural_cause)) +
  geom_histogram(binwidth = 500, fill = "skyblue", color = "black") +
  labs(
    x = "Natural Cause Mortality Count",
    y = "Frequency",
    title = "Distribution of Natural Cause Mortality"
  )
```

```{r}
#Synthetic data creation
obs <- nrow(data_filtered) #Number of observations

syn <- data.frame(
  
  #Set non-mortality data to match the previous dataset
  data_as_of = rep(as.Date('02/07/25', format = "%m/%d/%y"), obs),
  jurisdiction_of_occurance = rep('United States', obs),
  mmwr_year = c(rep(2020, nrow(filter(data_filtered, mmwr_year == 2020))),
                rep(2021, nrow(filter(data_filtered, mmwr_year == 2021))),
                rep(2022, nrow(filter(data_filtered, mmwr_year == 2022))),
                rep(2023, nrow(filter(data_filtered, mmwr_year == 2023)))),
  mmwr_week = c(1:nrow(filter(data_filtered, mmwr_year == 2020)),
                1:nrow(filter(data_filtered, mmwr_year == 2021)),
                1:nrow(filter(data_filtered, mmwr_year == 2022)),
                1:nrow(filter(data_filtered, mmwr_year == 2023))),
  
  #Simulate the mortality variables using normal distributions
  all_cause = rnorm(obs, mean = mean(data_filtered$all_cause), sd = sd(data_filtered$all_cause)),
  natural_cause = rnorm(obs, mean = mean(data_filtered$natural_cause), sd = sd(data_filtered$natural_cause)),
  cancer_count = rnorm(obs, mean = mean(data_filtered$cancer_count), sd = sd(data_filtered$cancer_count)),
  diabetes_count = rnorm(obs, mean = mean(data_filtered$diabetes_count), sd = sd(data_filtered$diabetes_count)),
  influenza_pneumonia_count = rnorm(obs, mean = mean(data_filtered$influenza_pneumonia_count), sd = sd(data_filtered$influenza_pneumonia_count)),
  heart_disease_count = rnorm(obs, mean = mean(data_filtered$heart_disease_count), sd = sd(data_filtered$heart_disease_count))
)
```

After synthesizing the new dataset, I will plot histograms of each of the mortality variables and compare to the original dataset.

```{r}
#Plotting distributions
#All-cause mortality
ggplot(syn, aes(x = all_cause)) +
  geom_histogram(binwidth = 500, fill = "skyblue", color = "black") +
  labs(
    x = "All Cause Mortality Count (Synthetic Data)",
    y = "Frequency",
    title = "Distribution of All Cause Mortality"
  )

#Natural cause mortality
ggplot(syn, aes(x = natural_cause)) +
  geom_histogram(binwidth = 500, fill = "skyblue", color = "black") +
  labs(
    x = "Natural Cause Mortality Count (Synthetic Data)",
    y = "Frequency",
    title = "Distribution of Natural Cause Mortality"
  )

#Cancer mortality
ggplot(syn, aes(x = cancer_count)) +
  geom_histogram(binwidth = 50, fill = "skyblue", color = "black") +
  labs(
    x = "Cancer Mortality Count (Synthetic Data)",
    y = "Frequency",
    title = "Distribution of Cancer Mortality"
  )

#Diabetes mortality
ggplot(syn, aes(x = diabetes_count)) +
  geom_histogram(binwidth = 50, fill = "skyblue", color = "black") +
  labs(
    x = "Diabetes Mortality Count (Synthetic Data)",
    y = "Frequency",
    title = "Distribution of Diabetes Mortality"
  )

#Influenza and pneumonia mortality
ggplot(syn, aes(x = influenza_pneumonia_count)) +
  geom_histogram(binwidth = 50, fill = "skyblue", color = "black") +
  labs(
    x = "Influenza and Pneumonia Mortality Count (Synthetic Data)",
    y = "Frequency",
    title = "Distribution of Influenza and Pneumonia Mortality"
  )

#Heart disease mortality
ggplot(syn, aes(x = heart_disease_count)) +
  geom_histogram(binwidth = 50, fill = "skyblue", color = "black") +
  labs(
    x = "Heart Disease Mortality Count (Synthetic Data)",
    y = "Frequency",
    title = "Distribution of Heart Disease Mortality"
  )
```

These distributions match the original data well in terms of their range of values and where the mean lies. However, there are some variations in the shape of the data. While the mortality data in the original dataset was largely normal, some outliers that were present in the original data were gave some of the distributions a slight skew. The simulations here incorporated those outlier values in the main distrubution, so the shape of the distributions differed in the synthetic data. 